A Handbook for Tenured Track Faculty in Computer Science in the US

### 2024-07-10 demystify.tex
updates

This section summarizes the main points of this guideline. This gives you an overview to decide which specific topics you want to explore more thoroughly.

\begin{enumerate}
  \item Should you apply?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries (\autoref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members (\autoref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---contact and introduce yourself (\autoref{sec:contact}).
         \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are letters of recommendation (LORs),  statements of purpose (SOP), and research experience and ability, e.g., publications.
          \item \emph{LORs are very important}, but only if they are personalized and talk about your research ability (\autoref{sec:lor}).
          \item \emph{SOP is very important}. Write it in such a way that makes you \emph{stand out} (\autoref{sec:research-statement} and \autoref{sec:improve-your-chance})
          \item  \emph{Publications can greatly help}. Publications in good venues are concrete evidence that you have successfully engaged in research. But they are not required! (\autoref{sec:research-experiences}).

          \item \emph{Grades might be important}, depends on the reputation of \emph{your school} (\autoref{sec:your-school}). While good grades might not help much, bad ones likely will hurt your case.

@@

          \item \emph{Contacting a prof. is recommended}, but do it \emph{properly} (\autoref{sec:contact}).

@@

        \end{itemize}
  \item What to do after getting admitted?
        \begin{itemize}
                \item  Celebrating! Now it is your turn to evaluate the school!
          \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\autoref{sec:accepted}).
        \end{itemize}
  \item Funding
        \begin{itemize}
          \item TAs, RAs, and fellowships are two main funding sources (\autoref{sec:funding}).  TAs (teaching assistantships) are provided by the department to help with classes (e.g., grading). RAs (research assistantships) are given by profs. to help with their research.  Fellowships, provided by the university, department, or external sources such as government or industry, give move flexibility but are very competitive.
        \end{itemize}
  \item Choosing School and Professors
        \begin{itemize}
          \item Many schools do not offer PhD studies in CS and many CS professors do not advise or graduate PhD students  (\autoref{sec:schoolsandprofs}).
        \end{itemize}

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item Being in an underrepresented group can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}) to apply for PhD in CS, and it takes about 5--7 years for PhD in CS (\autoref{sec:time}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipment and such for your research (\autoref{sec:buying-equipment}).
          %\item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}

@@

\paragraph{Acceptance Letters} Universities prioritize sending out acceptance letters first. This allows the admitted students to make decision and plan for their studies.

@@

\paragraph{Response Deadlines} Accepted students are usually given a deadline to make decision on their offers, often around \textbf{April 15th} in the US. This is a standard deadline for PhD programs due to the Council of Graduate Schools (CGS) Resolution (\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{see more}. After this date, CS programs can gauge how many slots remain unfilled.

@@

Given the high number of quality applicants and a limited number of spots, in addition to having a good application profile, you really want to show something that makes you \textbf{stand out}.  For example, even if you do not have research experience, you can talk about your personal projects, as long as they can help show you can do research. For example, if you have an open-source project that has lots of stars in Github, then you should mention it. If you often write technical, research-like blogs with many viewers, talk about them too.

@@

Finally, I make a decision, which is usually either a rejection or an offer of admission with full funding (e.g., from TA). I also recommend very strong candidates for the University Presidential fellowship. The adcom chair then reviews all evaluations, looking at entered notes, and ask reviewers to discuss and resolve discrepancies to reach a consensus (e.g., a reviewer wants to accept but the other wants to reject). Typically, the decision is made entirely by the reviewers; there is no involvement from the adcom chair, department chair, or others.

>>>>

This section summarizes the main points of this guideline. It also gives you an overview to decide which specific topics you want to explore more thoroughly.

\begin{enumerate}
  \item Should you apply to a PhD CS program in the US?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries (\autoref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members (\autoref{sec:evalapps}).
          \item Individually faculty member \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so do contact and introduce yourself (\autoref{sec:contact}).
         \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are letters of recommendation (LORs),  statements of purpose (SOP), and research experience and potential, e.g., publications.
          \item \emph{LORs are very important}, but only if they are personalized and talk about your research ability (\autoref{sec:lor}).
          \item \emph{SOP is very important}. Write it in such a way that makes you \emph{stand out} (\autoref{sec:research-statement} and \autoref{sec:improve-your-chance})
          \item  \emph{Publications can greatly help}. Papers in good venues are concrete evidence that you have successfully engaged in research. But they are not required! (\autoref{sec:research-experiences}).

          \item \emph{High grades probably won't help much} (unless you're from a well-known school), but bad ones likely will raise concerns (\autoref{sec:your-school}).

@@

          \item \emph{Contacting a prof. is recommended}, but do it \emph{properly} to get a reply (\autoref{sec:contact}).

@@

        \end{itemize}
  \item What to do after getting admitted?
        \begin{itemize}
                \item  Celebrating! Now it is your turn to evaluate the school!
          \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\autoref{sec:accepted}).
        \end{itemize}
  \item Funding
        \begin{itemize}
          \item TAs, RAs, and fellowships are main funding sources for PhDs (\autoref{sec:funding}).  TAs (teaching assistantships) are provided by the department to help with classes (e.g., grading). RAs (research assistantships) are given by profs. to help with their research.  Fellowships, provided by the university, department, or external sources such as government or industry, give move flexibility but are very competitive.
        \end{itemize}
  \item Choosing School and Professors
        \begin{itemize}
          \item Not every professor, even those in CS, can advise or graduate CS PhD students (\autoref{sec:schoolsandprofs}).
        \end{itemize}

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected}). One potential reason is because you're applying to AI/ML, a super competitive field in recent years with many applicants.
          \item Why rejection letters are sent so late? (\autoref{sec:late-rejection}). Grad programs often wait for the accepted students to make their decisions, typically by April 15, before sending out rejection letters.

          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item It take about 10 minutes to review an application (\autoref{sec:ievaluate}).
          \item Being a minority can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-CS background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}), and CS PhD takes about 5--7 years (\autoref{sec:time}). Compared to other countries, CS PhD in the US does not require an MS degree but has a longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call your professors? In many cases, just by first names! (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipment and such for your research (\autoref{sec:buying-equipment}).
          %\item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}

@@

\paragraph{Acceptance Letters} Universities prioritize sending out acceptance letters first. This allows the admitted students to make decisions and plan for their studies.

@@

\paragraph{Response Deadlines} Accepted students are usually given a deadline to make decisions on their offers, often around \textbf{April 15th} in the US. This is a standard deadline for PhD programs due to the Council of Graduate Schools (CGS) Resolution (\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{see more}. After this date, CS programs can gauge how many slots remain unfilled.

@@

Given the high number of quality applicants and a limited number of spots, in addition to having a good application profile, you want to show something that makes you \textbf{stand out}.  For example, even if you do not have research experience, you can talk about your personal projects, as long as they can help show you can do research. For example, if you have an open-source project that has lots of stars in Github, then you should mention it. If you often write technical, research-like blogs with many viewers, talk about them too.

@@

Finally, I make a decision, which is usually either a rejection or an offer of admission with full funding (e.g., from TA). I also recommend very strong candidates for the University Presidential fellowship. The adcom chair then reviews all evaluations, looks at entered notes, and asks reviewers to discuss and resolve discrepancies to reach a consensus (e.g., a reviewer wants to accept but the other wants to reject). Typically, the decision is made entirely by the reviewers; there is no involvement from the adcom chair, department chair, or others.

### 2024-07-10 faculty.tex
updates

\chapter{Research}
\chapter{Obtaining Funding}
Getting funding is arguably the most important part of getting tenure.  without funding, you can't pay your students, conference registration and travel, or even your own summer salary. However, writing grant proposals is very different from writing research papers. Thus, most new professors struggle with grant writing at first, but eventually get better at it.

@@

A strategy that I found effective is giving the entire class a small bonus point if everyone completes the evaluation. The bonus is computed as the percentage of responses * the bonus (X). So if X is 0.5 and everyone responds, everyone gets 0.5 points. If 50\% respond, each gets 0.25 points. While X is a tiny number, I emphasize that it can make a difference for borderline grades.  I don’t curve grades, so this is the only way for students to bump. I also emphasize that in my previous classes many students benefit from this.  So that really triggers their interest.  And also because evaluations are before final exams, everyone is incentivized to participate because they don’t know their final grades and don’t want to miss this opportunity.  Not only the students do it, they also remind their classmates/friends to do it.

I only employ this method recently, and response rate has been very good so far.  For Spr’24, it was 90\% (25/28 students) and for F’23, it was 100\% (16/16 students).  Before this, my response rate is about 50\%.

@@

\chapter{Services}

While only 10\% of your work is officially for services, they are very important for your tenure and career.
%You are expected to have certain types of services as a faculty member.  
Some services are required (e.g., when your department chair assigns you to some committee), but some are not and you can decline (e.g., an invitation to be a PC member).

@@

Common committees include PhD and/or MS admission, faculty search, and colloquium committees. Serving in the PhD adcom is very common for new assistant professors, as it is a good way to recruit students for your lab. Serving on the faculty search committee is also a good way to recruit new faculty members that you like (and also let you know what the job market is like, what is hot and not). Finally, being on the colloquium committee is a good way to invite speakers in your field that can be your \emph{tenure letter writers}.

You can also take more leadership roles when you are close to tenure, such as being the chair of a committee or chair/director of a program.  These are more time-consuming and might involve some politics, but they are also more rewarding and can help your tenure case.

@@

@@

Many would (rightly) advise about being strategic about your services. I do not really generally this and just volunteer for services that I think it is my responsibility (e.g., being in PCs, though I never liked them much) and those that I enjoy (NSF panels, services to the depts). In many cases I would also volunteer for tasks that no one wants to do because I feel that someone gotta do it and I can help out, e.g., various misc tasks in the dept.

>>>>

\chapter{Research}
\chapter{Obtaining Funding}
Getting funding is arguably the most important part of getting tenure.  Without funding, you can't pay your students, conference registration and travel, or even your own summer salary. However, writing grant proposals is very different from writing research papers. Thus, most new professors struggle with grant writing at first, but eventually get better at it.

@@

A strategy that I found effective is that the whole class gets a small bonus point if everyone completes the evaluation. The bonus is calculated as the percentage of responses multiplied by a factor (X). So, if X is 0.5 and everyone participates, they get 0.5 points each. If only half respond, it's 0.25 points each. It might seem small, but it can make a difference for borderline grades. Since I don't curve grades, this is their only chance to boost their scores. Many students have benefited from this in my previous classes, which really grabs their attention. Plus, since evaluations happen before finals and grades aren't finalized, everyone's motivated to participate—they don't want to miss out. Students not only do it themselves but also remind their friends. It's been working well for me: Spring '24 I had a 90\% response rate (25 out of 28 students), and Fall '23 was a perfect 100\% (16 out of 16 students). Before, my response rate was about 50\%, which is considered higher than average in the department.

@@

\chapter{Services}\label{sec:services}

Officially services are only counted for 10\% of your work, however they are very important for your tenure and career.
%You are expected to have certain types of services as a faculty member.  
Some services are required (e.g., when your department chair assigns you to some committees), but some are not and can be declined (e.g., an invitation to be a PC member).

@@

Common ones include PhD and/or MS admission, faculty search,  colloquium committees. Serving in the PhD adcom is very common for new assistant professors, as it is a good way to recruit students for your lab. Serving on the faculty search committee is also a good way to recruit new faculty members that you like (and also let you know what the job market is like, what is hot and not). Finally, being on the colloquium committee is a good way to invite speakers in your field that can be your \emph{tenure letter writers}.

Other types of services include leadership roles (e.g., being the director of a program), and representing the dept in various university committees.
These are not assigned by the dept, but rather you volunteer for them. Taking on a leadership role or representing the dept in certain university committees, which is less common for tenure-track profs., help the dept/chair in certain areas and can make you a good ``citizen'' which helps in your tenure case.

@@

\paragraph{PhD Dissertation Committee} Being in a PhD committee is a common task to help your colleagues (e.g., the adviser of the student asks you to be in student dissertation committee) and the students (e.g., you are involved or are interested in the research of the student).  You can serve in the committee of students in your dept, in other dept, or even in a different university (external committee member). This is a good way to build your network (and also help your students build their network).

@@

Many would (rightly) advise about being strategic about your services. I do not really generally this and just volunteer for services that I think it is my responsibility (e.g., being in PCs, though I never liked them much) and those that I enjoy (NSF panels, services to the depts). I enjoy services to my dept. and in many cases would take on tasks that no one wants to do because I feel that someone gotta do it and I can help out, e.g., various misc tasks in the dept.
